Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-07-31 paper/responseRound2/AJPResponseLetter.tex

Dear Joseph,

Thank you for the comments and feedback. 
We attach a new version of the manuscript.

We completely understand about the reference list and we have moved the majority of them into a ``Further Reading'' section in the Supplementary Material.

You also asked about whether we have had any experience using this apparatus in the classroom. Unfortunately we have not and we are unsure when this will be possible as we are in and out of lockdown in Melbourne. Before COVID, we used the interferometer at several science fairs and schools, however it did not have the audio set-up at that stage, so I think that experience is not directly relevant to this work.

@@

We have also made changes based on the annotated PDF from you and below answer the annotations where clarifications, etc. were needed.

Thank you again for the feedback.

Best wishes, \\
Hannah (for the co-authors)

\begin{itemize}
\item p.4 (Figure 4 comments):  I think it is obvious that a single FT of a frequency-varying signal cannot capture the signal, so I don't see the need for Figure 4. 
\begin{itemize}
\item We are happy to remove figure 4. We added it only after the initial review in response to one of the suggestions to add a comparison of what would be detected without using Viterbi. We agree, however, that it does not add much to the explanation. We have kept a shortened version of the highlighted text to explain why we cannot use the Fourier transform in this case and we have removed the figure. 
\end{itemize}

\item p.4 Section IV A: I appreciate your desire to be all-inclusive, but is the highlighted text necessary? It is not directly related to the topic of your paper. 
\begin{itemize}
\item This expanded literature review was one of the requests from Referee 3. We agree that they are tangentially related to the work presented, so we have instead moved them to the Further Reading section in the Supplementary Material. 
\end{itemize}

\item p.5 Figure 5: This won't show up in B/W. How about indicating the probability by different sizes of the circles? 
\begin{itemize}
\item We found it very difficult to make a tidy figure using different sized circles. Instead, we now use a grey colour scheme, which looks clear and should be fine in black and white print. 
\end{itemize}

\item p.5, left column, para 1: Just to make sure I understand: the color of each circle is what you observe?
\begin{itemize}
\item Yes, that's correct. We have added some clarification to the paragraph.
\end{itemize}

\item p.5, right column, para 2: The first two sentences seem to contradict each other. Can you clarify? 
\begin{itemize}
\item The first sentence refers to our prior beliefs before we begin any analysis or look at the data. Without any prior knowledge, we believe that the signal is equally likely to have started in any frequency bin within the range considered. The second sentence is the first stage of the algorithm once we have begun the analysis and looked at the data. We have added clarification to the text to explain this. 
\end{itemize}

\item p.5, right column, para 1: What does ``marginalization'' mean? Do you mean no hard barriers?
\item This statement was added in response to referee 3, who asked about marginalisation and spotted an error in this equation in the initial version of the manuscript. If we have a probability distribution in multiple dimensions, but we only want to consider the probability distribution of a single dimension, we would sum over (or marginalize) all of the other unwanted dimensions. This is not what we are doing in this instance. The equation corresponds to a single probability related to a unique path. We added this statement about marginalization to ensure no confusion given the referee's question, however we believe the meaning is still maintained without it. We have removed the first part of this sentence, keeping only the second part.

\item p.5, right column, list item 1: ``node'' meaning one of the circles in Fig. 5? 
\begin{itemize}
\item Yes, the nodes does mean the circles. Thank you, this was confusing and we've made the language more consistent now and referred to the circles when we first mention nodes.
\end{itemize}

\item p.5, right column, list item 3: In the example of Fig. 5, since A=1/3 or 0 for each transition, it looks like the selected path just links the states with the highest Fourier amplitude. Is it always that simple?
\begin{itemize}
\item That is a good point and it is not always so simple. In our Figure 5, we have made an example where the signal is particularly loud and can easily be seen by eye, but it is not always the case. In continuous-wave searches we expect that the signal cannot be seen by eye. It might also be the case that there are loud noise fluctuations in time-frequency bins. In this case they may not be selected by the Viterbi algorithm if the overall value of equation 5 is higher for the signal path than the paths containing the noise. 
\end{itemize}

\item p.6, left column, after itemized list: delete `expected to be' 
\begin{itemize}
\item We believe it is important to keep ``expected to be''. The reason is that continuous waves have not yet been detected and although of course we believe they are worth searching for, a more pessimistic person might argue that the signals may not be there at all! 
\end{itemize}

\item p.6, left column: How does the assumption of white noise get inserted into the VA? How does this assumption affect the results?
\begin{itemize}
\item The white noise assumption is implicit in that we do not build any knowledge of coloured noise into the analysis. This statement was added in response to referee 3's observation that the noise in our data is not white and they asked how or if we take account of this. We do not take account of non-white noise in our analysis (which is also consistent with LIGO analyses which assume white noise even though LIGO noise is not white). We also note that the narrow frequency range we consider is less impacted by the increase in noise at low frequency. We have added some details to clarify this in the text. 
\end{itemize}

\item p.6, right column: If you eliminate the start-up points from the calculation, does E change significantly?
\begin{itemize}
\item We have also computed the $E_{\rm rms}$ value when the first four time bins are ignored. We find a slight improvement in recovery and have added a sentence about this to the manuscript at the end of this paragraph. 
\end{itemize}

\item p.7, Fig. 6: It looks like the frequency bins are 0.3 Hz wide and 3s long. Can you estimate the S/N ratio in this data set to illustrate the power of the VA?
\begin{itemize}
\item We can only estimate S/N for single time blocks of data, however we do not believe that a representative S/N can be estimated for the entire data set. 
\end{itemize}

\item p.7, right column, Section V, para 2. ``Just to be clear, what have you changed from the apparatus in earlier sections (if anything), other than the photodiode described below? It reads like you have changed something.''
\begin{itemize}
\item The only thing we have changed is replacing the webcam with the photodiode so that the apparatus can be used to capture higher frequency speech and music. In this section we use the interferometer to capture more complicated signals, so we draw comparisons to laser microphones used in other fields. We have added a sentence to this paragraph to clarify the apparatus changes. 
\end{itemize}

\item p.8, left column: This paragraph adds very little new information to the topic of Section V, and interrupts the flow of the paper. I recommend shortening it. 
\begin{itemize}
\item Agreed. This paragraph was added in response to reviewer 3's request for us to mention other methods of noise reduction from the gravitational-wave literature. This seemed like the most appropriate place for it to go, however we agree that it is not closely related to the work presented here. We have shortened this discussion to a single sentence and moved the literature review to the Further Reading section in the Supplementary Material. 
\end{itemize}

\item p.9, Figure 8: It's hard to judge the effectiveness of the filtering on speech intelligibility from the graphs. Here's an idea: we could attach a direct link to an audio clip for each row, perhaps a 10 second recitation of a familiar phrase. It would be available to online readers via s simple click. This would enhance the discussion significantly.
\begin{itemize}
\item This sounds a great idea! We have collated the four sounds files and attached them in the email. Please let us know if you need any further details from us. 
\end{itemize}

\item p.10, left column, para 2: This seems like an unlikely approach. Why not simply increase the sensitivity, as described in the following sentence, and rotate the apparatus relative to the source?
\begin{itemize}
\item We have clarified this in the text. The gain in having data from two interferometer is that directional information can be estimated based on the time delay between the signal arriving at each interferometer, in a similar way to how we are able to tel the direction of a sound with our ears. This is also how gravitational-wave detectors estimate the direction of compact binary mergers based on the time-delay of the signal reaching two or more detectors. 
\end{itemize}

\item p.10, left column, para 2: Again, this seems like an unlikely strategy, given the sensitivity of the apparatus to mechanical noise. 
\begin{itemize}
\item We agree, this is likely to be a fair assessment about the noise of the system. We have removed the part about moving the interferometer. 
\end{itemize}

\item p.10, conclusions comments
\begin{itemize}
\item We have removed the repetition and shorted the conclusion. 
\end{itemize}

\item p.11, appendix B: ``The result you derive is a dc signal, which carries no information, plus a frequency-doubled component for small delta-d. But shouldn't there also be a component at the initial frequency, akin to frequency modulation? Otherwise, Fig. 3 cannot be explained.''
\begin{itemize}
\item Thank you for the careful reading and we agree completely - this was an error on our part in the derivation. We have decided to remove this appendix. 
\end{itemize}

\item p/11, appendix C: ``This needs more explanation. I certainly don't understand it.''
\begin{itemize}
\item We have expanded the text to clarify the explanation. 
\end{itemize}

### 2021-08-02 paper/ifo-tableTopGWs.tex
proof reading, minor typo fixes and edit

Sound is a commonly used analogy when explaining gravitational-wave science. 
Gravitational-wave signals from binary black hole and neutron star mergers can be converted to audio signals~\footnote{When introducing the acoustic analogy to lay audiences, it is important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot. Gravitational waves also travel at the speed of light and are not longitudinal waves.} to aid in explanations.
Detection and analysis techniques used by the gravitational-wave community can be demonstrated using table-top equipment (see also the further reading section in supplementary material for a selection of other table-top Michelson interferometer designs used for science communication).
Sound vibrations provide a simple means to move the components of a demonstration interferometer, changing the length of the interferometer arms, and therefore the interference pattern.
We use audio signals throughout this work to simulate gravitational-wave-like signals.

@@

The optical configuration of the Michelson interferometer used in this work is shown in Fig.~\ref{fig:ifo_schematic_webcam}.
It is assembled on a $450 \times 450 \,{\rm mm} $ optical breadboard and uses a green laser (top-left) with peak emission at a wavelength of $532\,{\rm nm}$.
The output of the laser is first reflected by a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard and to have greater control of the alignment of the interferometer).
After passing through a converging (plano-convex) lens with focal length $125.0\,{\rm mm}$, the laser beam is incident on a beamsplitter, which reflects half of the light towards mirror 1 (M1) and transmits the other half to mirror 2 (M2). 
A $0.5\,{\rm W}$ speaker is fixed to the back of M2 using commercial adhesive putty and serves as a controllable source of vibrations. 
This speaker is one of a pair of commercial, USB-powered speakers, fed by a $3.5\,{\rm mm}$ jack and driven by a personal computer (PC). 
The other speaker in the pair is kept face-down and as far away from the interferometer as possible to prevent it from acting as a second source.
The beams are reflected by mirrors M1 and M2, located at the end of $\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long arms.
The beams recombine at the beamsplitter and produce an interference pattern that is then enlarged using a diverging (bi-concave) lens of focal length $-25.0\,{\rm mm}$, and projected onto a screen.
The interference pattern, as shown by the photograph in Fig.~\ref{fig:interference_pattern}, is a set of concentric light and dark rings (fringes). 
A change in the relative length of the arms causes these rings to move radially in or outwards.
The intensity timeseries is recorded by webcam (Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}) or photodiode, which offers a higher sampling rate suitable for capturing more complex audio (Section~\ref{sec:optical_microphone}).
Further design details for Michelson interferometers can be found in Ref.~\cite{TTExhibit:2021}.

\begin{figure}
 \begin{center}
  \includegraphics[width=0.35\textwidth, angle=-90]{figures/webcam_still0_star.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
The interference pattern produced by the table-top Michelson interferometer.
This image was taken with the webcam aligned off the beam axis. %, which may explain why the (predicted circular) pattern appears elliptical.
The pink star indicates the point where the intensity timeseries measurements were taken.
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

The motion of the interference fringes follows the motion of M2, and therefore of the speaker.
The amplitude of these motions is given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the relative change in the length of the arms is kept small enough, then the intensity of the interference pattern at any point on the screen oscillates at the same frequency as the injected audio.
For larger relative length changes, multiple fringes will pass through the detection point during a single speaker oscillation, raising the measured frequency artificially. 
As such, any motion of the fringes must be kept small by playing the sound softly through the speaker.
% too much to have --- and () in the same sentence I think 
Even without over-counting, large fringe motions display a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting---phenomena like frequency doubling.

% removed as almost identical to the next sentence
%In the following section, we start simply by injecting and recovering an audio signal of constant frequency (a single tone).

% Data can be recorded via a webcam (as shown in Fig.~\ref{fig:ifo_schematic_webcam}) or photodiode. 
% In Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} a webcam is used. 
% In Section~\ref{sec:optical_microphone}, which considers more complex audio, we turn to a photodiode because it offers a higher sample rate than the webcam.

>>>>

Sound is a commonly used analogy when explaining gravitational-wave science. 
Gravitational-wave signals from binary black hole and neutron star mergers can be converted to audio signals~\footnote{When introducing the acoustic analogy to lay audiences, it is important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot. Gravitational waves also travel at the speed of light and are not longitudinal waves.} to aid in explanations.
Detection and analysis techniques used by the gravitational-wave community can be demonstrated using table-top equipment (see Ref.~\cite{TTExhibit:2021} as well as the further reading section in Supplementary Material for a selection of other table-top Michelson interferometer designs used for science communication).
Sound vibrations provide a simple means to move the components of a demonstration interferometer, changing the length of the interferometer arms, and therefore the interference pattern.
We use audio signals throughout this work to simulate gravitational-wave-like signals.

@@

The optical configuration of the Michelson interferometer used in this work is shown in Fig.~\ref{fig:ifo_schematic_webcam}.
It is assembled on a $450 \times 450 \,{\rm mm} $ optical breadboard and uses a green laser (top-left) with peak emission at a wavelength of $532\,{\rm nm}$.
The output of the laser is first reflected by a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard and to have greater control of the alignment of the interferometer).
After passing through a converging (plano-convex) lens with focal length $125.0\,{\rm mm}$, the laser beam is incident on a beamsplitter, which reflects half of the light towards mirror 1 (M1) and transmits the other half to mirror 2 (M2). 
A $0.5\,{\rm W}$ speaker is fixed to the back of M2 using commercial adhesive putty and serves as a controllable source of vibrations. 
This speaker is one of a pair of commercial, USB-powered speakers, fed by a $3.5\,{\rm mm}$ jack and driven by a personal computer (PC). 
The other speaker in the pair is kept face-down and as far away from the interferometer as possible to prevent it from acting as a second source.
The beams are reflected by mirrors M1 and M2, located at the end of $\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long arms.
The beams recombine at the beamsplitter and produce an interference pattern that is then enlarged using a diverging (bi-concave) lens of focal length $-25.0\,{\rm mm}$, and projected onto a screen.
The interference pattern, as shown by the photograph in Fig.~\ref{fig:interference_pattern}, is a set of concentric light and dark rings (fringes). 
A change in the relative length of the arms causes these rings to move radially in or outwards.
The intensity timeseries is recorded by either a webcam (in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}) or a photodiode, which offers a higher sampling rate suitable for capturing more complex audio (in Section~\ref{sec:optical_microphone}).
Further design details for Michelson interferometers can be found in Ref.~\cite{TTExhibit:2021}.

\begin{figure}
 \begin{center}
  \includegraphics[width=0.35\textwidth, angle=-90]{figures/webcam_still0_star.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
The interference pattern produced by the table-top Michelson interferometer.
This image was taken with the webcam aligned off the beam axis. 
The pink star indicates the point where the intensity timeseries measurements were taken.
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

The motion of the interference fringes follows the motion of M2, and therefore of the speaker.
The amplitude of these motions is given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the relative change in the length of the arms is kept small enough, then the intensity of the interference pattern at any point on the screen oscillates at the same frequency as the injected audio.
For larger relative length changes, multiple fringes will pass through the detection point during a single speaker oscillation, raising the measured frequency artificially. 
As such, any motion of the fringes must be kept small by playing the sound softly through the speaker.
Even without over-counting, large fringe motions display a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting---phenomena like frequency doubling.

### 2021-08-02 paper/ifo-wanderingTone.tex
proof reading, minor typo fixes and edit

%%%%%%%%%%%%%%%%%%%%%% Comment from editor to remove this figure
%We first demonstrate why a Fourier transform applied to the whole dataset (as in Section~\ref{sec:single_tone}) is not well suited to the wandering frequency case.
%As shown in Fig.~\ref{fig:viterbi_comparison_webcam_spectrum}, the signal is spread across multiple frequency bins, making it challenging or in some cases impossible to detect.

A Fourier transform applied to the whole dataset (as in Section~\ref{sec:single_tone}) is not well suited to the wandering frequency case as the signal is spread across multiple frequency bin. 
In continuous wave searches, the wandering frequency problem is solved using the Viterbi algorithm~\cite{Viterbi:1967}, which can track the signal's frequency over time.
The analysis described and presented in this section is directly inspired by real continuous-wave search methods~\cite{SuvorovaEtAl:2017}. 
In Section~\ref{sec:realCWSearches}, we briefly review the analysis method used by LIGO and Virgo; in Section~\ref{sec:viterbi}, we describe the methods as applied in this work; and in Section~\ref{sec:wanderingResults}, we show results for recovering a wandering frequency signal using the table-top interferometer.

% Removing figure after editor request. 
\begin{comment}
\begin{figure}
	\includegraphics[width=.5\textwidth]{figures/viterbi_comparison_entire_spectrum_viterbi_test_webcam.pdf}
	\caption{\label{fig:viterbi_comparison_webcam_spectrum}
Fourier amplitude identical to that shown in Fig.~\ref{fig:webcam_spectrum}, except using a signal which has a slowly changing frequency instead of a constant frequency signal.
The signal is spread between many frequency bins. 
We see peaks at $5.7\,{\rm Hz}$ and $10.7\,{\rm Hz}$ and refer the reader forward to Fig.~\ref{fig:viterbi_overlay} to see the frequency evolution of this signal. 
A single Fourier transform of the entire dataset is not well suited to detecting signals with wandering frequencies.
% just making this a little more concise 
%Why we use the Viterbi algorithm to recover wandering frequency signals: Here we plot the Fourier amplitude of the intensity pattern against frequency as in Fig.~\ref{fig:webcam_spectrum} for the entire timeseries of the wandering frequency signal in Fig.~\ref{fig:viterbi_overlay}. Instead of resolving definite peaks like the constant frequency signal shown in Fig.~\ref{fig:webcam_spectrum}, we only see inconclusive peaks at 5.7~Hz and 10.7~Hz that correspond to the long amount of time that the wandering frequency spends near those values and/or its amplitude there, as seen in the pink dots in Fig.~\ref{fig:viterbi_overlay}. With the signal spread out over the frequency bins, we are unable to reliably detect it above the noise and therefore we use the Viterbi algorithm instead.
%The large values at low frequencies below 0.1~Hz are similar to those in Fig.~\ref{fig:webcam_spectrum} and can be ignored as they are likely inaccurate due to the short length of the signal relative to their period (i.e.\ 300~s only contains 30~periods of a 0.1~Hz signal), but this feature would require further research to fully determine its origin. %maybe could be from the long period (100~s) variation?
}
\end{figure}
\end{comment}

@@

%\han{from Andrew: sec IVA, first para - since ref mentioned other searches, it would be good to have 1-2 brief sentences accompanying the references which mention what those "other searches" are (broad categories, for instance, just so that there is some guidance for the reader)}

The methods used here are inspired by LIGO-Virgo continuous wave searches. 
For further details on these methods and continuous wave searches generally we refer the reader to references in the Supplementary Material.

Continuous-wave searches are performed on long datasets, months to years in duration. 
The frequency of the signal can wander significantly over the observation period. 
In this context, ``significantly'' means across multiple frequency bins, where the typical width of a frequency bin is the reciprocal of the total observation time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.
%One method to search for a wandering signal is to split the timeseries data into several shorter segments which are analyzed individually.

@@

%Before beginning the analysis, we assume that the signal is equally likely to start in any frequency bin (i.e. the prior is flat). 
%The first step of the analysis is to define the probability of the system having frequency $f_j$ at

Before we begin analysing the data, we have no prior knowledge as to which frequency bin the signal starts from (i.e. the prior is flat between the minimum and maximum frequency bins).

@@

%We note that there is no marginalization the left-hand-side and right-hand-side of Eqn.~\ref{eqn:pathprobability} are both evaluated for a specific path and then we maximize over all such paths to find the optimal Viterbi path. 
%The probability of the system having frequency $f_j$ at any later time $t_{i+1}$ is then defined recursively by
%\begin{eqnarray}
%{\rm Pr}[f(t_{i+1})=f_j] =~& F[t_{i+1},f(t_{i+1})] \nonumber \\
%                     &\times A[f(t_{i+1}),f(t_i)]  \nonumber \\
%                     &\times {\rm Pr}[f(t_i)].
%\label{eqn:recursiveViterbi}
%\end{eqnarray}
%for any path through the grid. 
%The sequence $f(t_0),f(t_1),\dots,f(t_{N_t})$ that maximizes ${\rm Pr}[f(t_{N_t}) = f_j]$ is the optimal Viterbi path terminating in the frequency bin $f_j$. 
%We can then maximize the latter quantity over $0 \leq j \leq N_f$ to find the optimal Viterbi path overall, i.e. terminating in any frequency bin.

The Viterbi algorithm provides a computationally efficient method for finding the optimal path. 
At every $t_i$ all but $N_f$ possible paths are eliminated. 
Here we describe the algorithm in pseudocode while referring to the schematic in Fig.~\ref{fig:viterbi}. %, where the circles indicate the elements in the spectrogram with the lighter colors corresponding to a higher likelihood. 
The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at time $t_1$, each $f_j$ node (circles in Fig~\ref{fig:viterbi}) can originate from three prior nodes at time $t_0$ (except for the edge cases $f_0$ and $f_{N_f}$ which only have two). 
The paths between these nodes are indicated by the lines in Fig.~\ref{fig:viterbi}. 
At each $f_j$ node, we select the path with the highest $A[f(t_0),f(t_1)] {\rm Pr}[f(t_0)]$ value as the most probable path. 
These choices are highlighted using the black lines in Fig.~\ref{fig:viterbi} while the gray lines show the rejected paths. 
For example, the most probable connection to the node labeled (a) is the one directly behind it (i.e., $f_4$). 
Therefore, this path is selected as the best path from $t_0$ to $t_1$ for $f_4$.
To allow backtracking at the end, the index of the most probable connection to a node along with the value of the best path to that node is stored, for each node.

@@

\item The final step is to find the Viterbi path (the overall best path that terminates in the frequency bin with the highest probability in step 4). 
The Viterbi path is found by backtracking along the stored best connections at each $t_i$ (see also Appendix~\ref{app:viterbi}). 
In Fig.~\ref{fig:viterbi}, it is the path ending at (b) that started at (c) highlighted in orange.
%\item The Viterbi path (the overall best path) is then found by backtracking along the stored best connections at each $t_i$ (see also Appendix~\ref{app:viterbi}). In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c) and is highlighted in orange.
\end{enumerate}

In continuous gravitational wave searches, the signal amplitude is expected to be small in comparison to the noise and its frequency can change unpredictably over time.
The Viterbi algorithm's strength lies in its ability to track such signals through the data.
In the following section, we present the results of using the Viterbi algorithm with the table-top interferometer data. 
% - just simiplifed this discussion a little and added a reference to the CW searches
%The key to the Viterbi algorithm is that we can use the signal's correlation in time to improve the signal-to-noise ratio of the frequency of the wandering signal at a particular time. Since the noise is assumed to not be correlated in time, if the detection statistic at a given frequency is high at a particular time and at neighboring frequencies at neighboring times then the likelihood that the given frequency at that particular time is from the wandering signal increases. By iterating this process through the grid for a suitable choice of the window of neighboring frequencies and times, the Viterbi algorithm is well-suited to finding wandering signals.

% The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
% The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
% The algorithm is computationally efficient, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}).

@@

This may be explained by two anomalies in the recovered path. Initially, the injected signal wanders by more than one frequency bin per time bin (i.e.\ faster than our choice of $A(f_k,f_m)$ allows), thus leading to a discrepancy between the injected and recovered paths for $t\lesssim 10\,{\rm s}$. One may be tempted to increase the allowed frequency wander in the algorithm, however, this leads to an overall decrease in the above statistics, as the algorithm is prone to jump briefly to nearby spots of noise. There is also an anomaly at $150\,{\rm s}$, which is likely due to a local disturbance, e.g. someone walking past the interferometer. 
As shown in Fig.~\ref{fig:viterbi_overlay}, the Viterbi algorithm can recover and continues to track the signal after the disturbance. 
%The Viterbi algorithm is somewhat robust against such disturbances, qualitatively recovering the injected path on the other side, as shown in Fig.~\ref{fig:viterbi_overlay}.

>>>>

A Fourier transform applied to the whole dataset (as in Section~\ref{sec:single_tone}) is not well suited to the wandering frequency case as the signal is spread across multiple frequency bins. 
In continuous-wave searches, the wandering frequency problem is solved using the Viterbi algorithm~\cite{Viterbi:1967}, which can track the signal's frequency over time.
The analysis described and presented in this section is directly inspired by real continuous-wave search methods~\cite{SuvorovaEtAl:2017}. 
In Section~\ref{sec:realCWSearches}, we briefly review the analysis method used by LIGO and Virgo; in Section~\ref{sec:viterbi}, we describe the methods as applied in this work; and in Section~\ref{sec:wanderingResults}, we show results for recovering a wandering frequency signal using the table-top interferometer.

@@

The methods used here are inspired by LIGO and Virgo continuous-wave searches. 
For further details on these methods and continuous-wave searches generally we refer the reader to references in the Supplementary Material.

Continuous-wave searches are performed on long datasets, months to years in duration. 
The frequency of the signal can wander significantly over the observation period. 
In this context, ``significantly'' means across multiple frequency bins, where the typical width of a frequency bin is the reciprocal of the total observation time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.

@@

Before we begin analyzing the data, we have no prior knowledge as to which frequency bin the signal starts from (i.e. the prior is flat between the minimum and maximum frequency bins).

@@

The Viterbi algorithm provides a computationally efficient method for finding the optimal path. 
At every $t_i$ all but $N_f$ possible paths are eliminated. 
Here we describe the algorithm in pseudocode while referring to the schematic in Fig.~\ref{fig:viterbi}.  
The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at time $t_1$, each $f_j$ node (circles in Fig~\ref{fig:viterbi}) can originate from three prior nodes at time $t_0$ (except for the edge cases $f_0$ and $f_{N_f}$ which only have two). 
The paths between these nodes are indicated by the lines in Fig.~\ref{fig:viterbi}. 
At each $f_j$ node, we select the path with the highest $A[f(t_0),f(t_1)] {\rm Pr}[f(t_0)]$ value as the most probable path. 
These choices are highlighted using the black lines in Fig.~\ref{fig:viterbi} while the gray lines show the rejected paths. 
For example, the most probable connection to the node labeled (a) is the one directly behind it (i.e., $f_4$). 
Therefore, this path is selected as the best path from $t_0$ to $t_1$ for $f_4$.
To allow backtracking at the end, the index of the most probable connection to a node along with the value of the best path to that node is stored, for each node.

@@

\item The final step is to find the Viterbi path (the overall best path that terminates in the frequency bin with the highest probability in step 4). 
The Viterbi path is found by backtracking along the stored best connections at each $t_i$ (see also Appendix~\ref{app:viterbi}). 
In Fig.~\ref{fig:viterbi}, it is the path ending at (b) that started at (c) highlighted in orange.
\end{enumerate}

In continuous gravitational wave searches, the signal amplitude is expected to be small in comparison to the noise and its frequency can change unpredictably over time.
The Viterbi algorithm's strength lies in its ability to track such signals through the data.
In the following section, we present the results of using the Viterbi algorithm with the table-top interferometer data.

@@

This may be explained by two anomalies in the recovered path. Initially, the injected signal wanders by more than one frequency bin per time bin (i.e.\ faster than our choice of $A(f_k,f_m)$ allows), thus leading to a discrepancy between the injected and recovered paths for $t\lesssim 10\,{\rm s}$. One may be tempted to increase the allowed frequency wander in the algorithm, however, this leads to an overall decrease in the above statistics, as the algorithm is prone to jump briefly to nearby spots of noise. There is also an anomaly at $150\,{\rm s}$, which is likely due to a local disturbance, e.g. someone walking past the interferometer. 
As shown in Fig.~\ref{fig:viterbi_overlay}, the Viterbi algorithm can recover and continues to track the signal after the disturbance.
